---
output:
  pdf_document: default
  html_document: default
---
# Reassessing the evidence for universal school-age Bacillus Calmette Guerin (BCG) vaccination in England and Wales {#sutherland}


```{r 05-knitr-options, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi = 320,
                      fig.width = 8, fig.height = 8,
                      out.width = "80%", fig.align = 'center')
```

```{r 05-setup, include = FALSE}
## Packages
library(tidyverse)
library(prettypublisher)
library(knitr)
library(kableExtra)

## Read formated results
## See https://github.com/seabbs/AssessBCGPolicyChange/blob/master/vignettes/paper.Rmd
## for code required to reproduce 
## Chapter is an edited version of the orginal reproducible document.
##Wordcount: 


## Functions for results read in
read_formated <- function(name) {
  path <- file.path("chapters/evidence-policy-change", paste0(name, ".rds"))
  
  tmp <- readRDS(path)
  
  return(tmp)
}


load_formated <- function(name) {

  path <- file.path("chapters/evidence-policy-change", paste0(name, ".rda"))
  
  load(path)
}

## Read in all relevent rda's
rda_list <- c("annual_nots-mw", "annual_nots_tab", "avg-inc", "avg-not",
              "mann-whitney", "max-not",
              "min-not", "model-valid",
              "table-vac-prev")

walk(rda_list, ~ load_formated(.))


```

## Introduction


Prior to the change in BCG vaccination policy in 2005 (see Chapter \@ref(background)) several studies were carried out to assess the impact of any potential policy change. In this Chapter I aimed to update one of these studies.

I recreated a previous approach for estimating the impact of ending the BCG schools scheme in England and Wales, updating the model with parameter uncertainty, and measurement error. I investigated scenarios considered by the Joint Committee on Vaccination and Immunisation, and explored new approaches using notification data (see Chapter \@ref(data)). I estimated the number of vaccines needed to prevent a single notification, and the average annual additional notifications caused by ending the BCG schools’ scheme. This work was supervised by Hannah Christensen and Ellen Brooks-Pollock.









## Summary


* I found a 1.9% annual decrease in Tuberculosis incidence rates best matched our estimates from notifications.

* I estimate that 1600 (95% Interquartile range (IQR) 1300 to 2100) vaccines would have been required to prevent a single notification in 2004. 

* If the scheme had ended in 2001, 302 (95% IQR 238 to 369) additional annual notifications would have occurred compared to if the scheme had continued. If the scheme ended in 2016, 120 (95% IQR 88 to 155) additional annual notifications would have occurred.

* The estimates of the impact of ending the BCG schools scheme were highly sensitive to the annual decrease in incidence rates. 

* The impact of ending the BCG schools scheme was found to be greater than previously thought when parameter values were updated and notifications were used.

* The results highlight the importance of including uncertainty when forecasting the impact of changes in vaccination policy.

* The code for the analysis contained in this Chapter can be found at: [doi.org/10.5281/zenodo.2635687](https://doi.org/10.5281/zenodo.2635687)^[Alternatively available from: https://github.com/seabbs/AssessBCGPolicyChange]
