---
title: "Thesis timeline"
author: "Sam Abbott"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
```

## Overview of Timeline

This timeline gives a breakdown on the current status of each PhD chapter. It then outlines the monthly targets towards a completion date of the 01/05/19.

## Chapter Breakdown

1. **Introduction:**
    - Status: Planned and partially written
    
1. **The epidemiology of tuberculosis, and the role of BCG vaccination, in England:** 
    - Status: Draft complete, needs updating based on comments from Matt.
    
1. **Reassessing the Evidence for Universal School-age Bacillus Calmette Guerin (BCG) Vaccination in England and Wales:**
    - Status: Complete as draft paper - requires some updating based on work from the JVCI and conversion into thesis chapter
    
1. **Beneficial effects of BCG vaccination in outcomes for patients diagnosed with TB: observation study using the Enhanced Tuberculosis Surveillance System 2009-2015:**
    - Status: Under review at Vaccine (5 months) - requires conversion
    
1. **Estimating the effect of the 2005 UK BCG vaccination policy change: A retrospective cohort study using the Enhanced Tuberculosis Surveillance system, 2000-2015:**
    - Status: Complete as draft paper - requires conversion
    
1. **Developing and parameterising a dynamic transmission model of tuberculosis:**
    - Status: Chapter complete but requires updating based on changes during model implementation and fitting


1. **Investigating the impact of the 2005 change in BCG vaccination policy using a fitted dynamic transmission model of TB**
    - Status: Model, model fitting pipeline, model scenario analysis and model sensitivity implemented. Preliminary results have been used to construct code to produce final plots/tables/output metrics. Run and validating model fitting is in progress. Chapter has been outlined and partially written up (~30%).

1. **Discussion:**
    - Status: Outlined.

## Monthly Breakdown

```{r, echo = FALSE, results = "asis"}
timeline_tab <- data_frame(
  Month = c("Feburary", ""),
            Sam = c(
              "1st Introduction",
              "15th, The epidemiology of tuberculosis, and the role of BCG vaccination, in England & 
              Developing and parameterising a dynamic transmission model of tuberculosis"),
            Supervisors = c("15th Introduction", "")
            ) %>% 
    add_row(Month = c("March", ""),
          Sam = c("1st Reassessing the Evidence for Universal School-age Bacillus Calmette Guerin (BCG) Vaccination in England and Wales & 
                  Beneficial effects of BCG vaccination in outcomes for patients diagnosed with TB: observation study using the Enhanced Tuberculosis Surveillance System 2009-2015",
                  "15th Investigating the impact of the 2005 change in BCG vaccination policy using a fitted dynamic transmission model of TB & 
                   Estimating the effect of the 2005 UK BCG vaccination policy change: A retrospective cohort study using the Enhanced Tuberculosis Surveillance system, 2000-2015"),
          Supervisors = c("1st The epidemiology of tuberculosis, and the role of BCG vaccination, in England & 
              Developing and parameterising a dynamic transmission model of tuberculosis",
              "")) %>% 
    add_row(Month = c("April"),
          Sam = c("1st All chapters excepting the last two chapters with Supervisors,
                  15th Final two chapters updated based on supervisor comments."),
          Supervisors = c("1st Investigating the impact of the 2005 change in BCG vaccination policy using a fitted dynamic transmission model of TB & 
                    Estimating the effect of the 2005 UK BCG vaccination policy change: A retrospective cohort study using the Enhanced Tuberculosis Surveillance system, 2000-2015")
          ) %>% 
    add_row(Month = c("May"),
          Sam = c("1st Final hand in + viva organisation"),
          Supervisors = c("")) 


kable(timeline_tab, booktabs = TRUE) %>% 
  column_spec(c(1), width = "2cm") %>% 
  column_spec(c(2, 3), width = "8cm") %>%
  landscape()
```





